A general Nullstellensatz for generalized spaces

### 2019-03-05 paper-qcoh.tex
More work on the introduction

A fundamental result is that there is a \emph{generic model}~$U_\TT$ of~$\TT$,
a model such that for any geometric sequent~$\sigma$, the following notions
coincide: \begin{enumerate} \item The sequent~$\sigma$ is provable
modulo~$\TT$. \item The sequent~$\sigma$ holds for any~$\TT$-model in any
Grothendieck topos. \item The sequent~$\sigma$ holds for~$U_\TT$.
\end{enumerate} One could argue that it is this model which mathematicians
implicitly refer to when they utter ``Let~$M$ be a~$\TT$-model.''. It can
typically not be realized as a set-theoretic model, consisting of a set for
each sort, a function for each function symbol and so on; instead it is a model
in a custom-tailored syntactically constructed Grothendieck topos, the
\emph{classifying topos}~$\Set[\TT]$ of~$\TT$, hence consists of an object
of~$\Set[\TT]$ for each sort, a morphism for each function symbol and so on.

@@

From the point of view of the topos, the ring object~$\affl$ is not only a
local ring, but even a field in the sense that
\[ \text{any nonzero element is invertible}. \]
As this condition is of nongeometric form, it is not inherited by arbitrary
local rings, which are indeed typically not fields. However, any intuitionistic
consequence of this condition which is of geometric form is inherited
by any local ring in any topos. Hence we may, when verifying a general fact
about local rings, suppose without loss of generality that the given ring is a
field. This observation is due to Kock~\cite{kock:univ-proj-geometry}, who exploited it to
develop projective geometry over local rings, and was further used by Reyes to
prove a Jacobian criterion for étale morphisms~\cite{reyes:cramer}.

@@

In time, further nongeometric sequents holding in the big Zariski topos of an
arbitrary base scheme have been found~\cite[Section~18.4]{blechschmidt:phd}. These include:
\begin{itemize}
\item $\affl$ is \emph{anonymously algebraically closed} in the sense that
any monic polynomial~$p \? \affl[T]$ of degree at least one does \notnot have a
zero.
\item $\affl$ fulfills the following version of the Nullstellensatz:
Let $f_1,\ldots,f_m \in \affl[X_1,\ldots,X_n]$ be polynomials without a common
zero in~$(\affl)^n$. Then there are polynomials~$g_1,\ldots,g_m \in
\affl[X_1,\ldots,X_n]$ such that~$\sum_i g_i f_i = 1$.
\item Any function~$\affl \to \affl$ is given by a unique polynomial.
\item $\affl$ fulfills the axiom of microaffinity: Let~$\Delta = \{ \varepsilon \?
\affl \,|\, \varepsilon^2 = 0 \}$. Let~$f : \Delta \to \affl$ be an arbitrary
function. Then there are unique elements~$a, b \? \affl$ such
that~$f(\varepsilon) = a + b \varepsilon$ for all~$\varepsilon \? \Delta$.
\item $\affl$ is \emph{synthetically quasicoherent} in the sense that for any
finitely presented~$\affl$-algebra~$A$, the canonical homomorphism~$A \to
(\affl)^{\Spec(A)}$, where~$\Spec(A)$ is defined to be the set
of~$\affl$-algebra-homomorphisms~$A \to \affl$, is bijective.
\end{itemize}
All of these nongeometric sequents are useful for the purposes of synthetic
algebraic geometry, the desire to carry out algebraic geometry in a language
close to the simple language on the 19th and the beginning of the 20th century
while still being fully rigorous and fully general, working over arbitrary base
schemes instead of restricting to the field of complex numbers.
\medskip

@@

\paragraph{A general Nullstellensatz} To explain the relevant background, the
somewhat vague question ``to which extent does the classifying
topos~$\Set[\TT]$ realize that it is the classifying topos for~$\TT$?'' is a
useful guiding principle.

From the point of view of~$\Set[\TT]$, the model~$U_\TT$ is an ordinary
set-theoretic model. While in general properties of a specific set-theoretic
model do not follow solely from the axioms of the theory, it is on first sight
conceivable that this could be true for~$U_\TT$. That it, is is reasonable to
wonder whether
\[ \Set[\TT] \models \speak{a geometric sequent~$\sigma$ holds for~$U_\TT$ iff~$\ul{\TT}$ proves~$\sigma$} \]
holds.\footnote{Here~$\ul{\TT}$ is the internal geometric theory induced
by~$\TT$, obtained by pulling back the set of sorts, the set of function
symbols and so on along the geometric morphism~$\Set[\TT] \to \Set$. More
details will be given in Section~\ref{sect:review}. The corner quotes indicate
that for sake of readability, the translation into formal language is to be
carried out by the reader.} While the ``if'' direction
certainly holds (observing that the generic model~$U_\TT$ is, from the point of
view of~$\Set[\TT]$, a model of~$\ul{\TT}$), the ``only if'' direction
typically fails.

@@

Intuitively, the problem is that while the meaning of~$\speak{$\ul{\TT} \vdash
\sigma$}$ is fixed, the meaning of~$\speak{$\sigma$ holds for~$U_\TT$}$ can vary
with the slice. This problem can be solved by passing from~$\ul{\TT}$ to a
varying theory, the internal theory~$\ul{\TT}/U_\TT$ defined below.
Unlike~$\ul{\TT}$, this theory is not the pullback of an external geometric
theory.

From the point of view of~$\Set[\TT]$, a model of~$\ul{\TT}/U_\TT$ is a model
of~$\ul{\TT}$ equipped with a~$\ul{\TT}$-homomorphism from~$U_\TT$. In
particular, the identity~$(U_\TT \to U_\TT)$ is a model of~$\ul{\TT}/U_\TT$.

\begin{thm}Let~$\TT$ be a geometric theory. Then, internally to~$\Set[\TT]$:
\begin{quote}
A geometric sequent~$\sigma$ over the signature of~$\ul{\TT}/U_\TT$ holds
for~$U_\TT$ if and only if~$\ul{\TT}/U_\TT$ proves~$\sigma$.
\end{quote}
\end{thm}

\begin{thm}Let~$\TT$ be a geometric theory. Let~$\alpha$ be a first-order
formula over the signature of~$\TT$. Then the following statements are
equivalent.
\begin{enumerate}
\item The formula~$\alpha$ holds for~$U_\TT$.
\item The formula~$\alpha$ is provable in first-order intuitionistic logic
modulo the axioms of~$\TT$ and the additional axioms
\[ \sigma \quad\Longleftrightarrow\quad
  \ul{\TT}/U \vdash' \sigma \]
where~$\sigma$ ranges over all geometric sequents.
\end{enumerate}
\end{thm}

@@

>>>>

A fundamental result is that there is a \emph{generic model}~$U_\TT$ of~$\TT$,
a model such that for any geometric sequent~$\sigma$, the following notions
coincide: \begin{enumerate} \item The sequent~$\sigma$ is provable
modulo~$\TT$. \item The sequent~$\sigma$ holds for any~$\TT$-model in any
Grothendieck topos. \item The sequent~$\sigma$ holds for~$U_\TT$.
\end{enumerate} One could argue that it is this model which mathematicians
implicitly refer to when they utter the phrase ``Let~$M$ be a~$\TT$-model.''. It can
typically not be realized as a set-theoretic model, consisting of a set for
each sort, a function for each function symbol and so on; instead it is a model
in a custom-tailored syntactically constructed Grothendieck topos, the
\emph{classifying topos}~$\Set[\TT]$ of~$\TT$, hence consists of an object
of~$\Set[\TT]$ for each sort, a morphism for each function symbol and so on.

@@

From the point of view of the topos, the ring object~$\affl$ is not only a
local ring, but even a field in the sense that
any nonzero element is invertible.
As this condition is of nongeometric form, it is not inherited by arbitrary
local rings, which are indeed typically not fields. However, any intuitionistic
consequence of this condition which is of geometric form is inherited
by any local ring in any topos. Hence we may, when verifying a general fact
about local rings, suppose without loss of generality that the given ring is a
field. This observation is due to Kock~\cite{kock:univ-proj-geometry}, who exploited it to
develop projective geometry over local rings, and was further used by Reyes to
prove a Jacobian criterion for étale morphisms~\cite{reyes:cramer}.

@@

In time, further nongeometric sequents holding in the big Zariski topos of an
arbitrary base scheme have been found~\cite[Section~18.4]{blechschmidt:phd}. These include:
\begin{itemize}
\item $\affl$ is \emph{anonymously algebraically closed} in the sense that
any monic polynomial~$p \? \affl[T]$ of degree at least one does \notnot have a
zero. \smallskip
\item The Nullstellensatz holds:
Let $f_1,\ldots,f_m \in \affl[X_1,\ldots,X_n]$ be polynomials without a common
zero in~$(\affl)^n$. Then there are polynomials~$g_1,\ldots,g_m \in
\affl[X_1,\ldots,X_n]$ such that~$\sum_i g_i f_i = 1$. \smallskip
\item Any function~$\affl \to \affl$ is given by a unique polynomial. \smallskip
\item $\affl$ is microaffine: Let~$\Delta = \{ \varepsilon \?
\affl \,|\, \varepsilon^2 = 0 \}$. Let~$f : \Delta \to \affl$ be an arbitrary
function. Then there are unique elements~$a, b \? \affl$ such
that~$f(\varepsilon) = a + b \varepsilon$ for all~$\varepsilon \? \Delta$. \smallskip
\item $\affl$ is \emph{synthetically quasicoherent}: For any
finitely presented~$\affl$-algebra~$A$, the canonical homomorphism~$A \to
(\affl)^{\Spec(A)}$, where~$\Spec(A)$ is defined as the set
of~$\affl$-algebra-homomorphisms~$A \to \affl$, is bijective.
\end{itemize}
All of these nongeometric sequents are useful for the purposes of synthetic
algebraic geometry, the desire to carry out algebraic geometry in a language
close to the simple language on the 19th and the beginning of the 20th century
while still being fully rigorous and fully general, working over arbitrary base
schemes instead of restricting to the field of complex numbers.
\medskip

@@

\paragraph{A general Nullstellensatz} To explain the relevant background, the
somewhat vague question ``to which extent does the classifying
topos~$\Set[\TT]$ realize that it is the classifying topos for~$\TT$?'' is a
useful guiding principle. This is easiest to visualize with a concrete example
for~$\TT$, such as the theory of rings.

Let~$A$ be a ring. A simple version of the classical Nullstellensatz states:
\emph{For any polynomials~$f$ and~$g$ over~$A$, if any zero of~$f$ is also a
zero of~$g$, then there is a polynomial~$h$ such that~$g = hf$.} The
polynomial~$h$ can be regarded as an ``algebraic certificate'' of the
hypothesis. This principle holds for instance in the case that~$A$ is an
algebraically closed field and that~$g$ is the unit polynomial. It is also
true, without any restriction on~$g$, for the generic ring.

We could try to generalize the Nullstellensatz to arbitrary geometric theories~$\TT$ as
follows. \emph{For any geometric sequent~$\sigma$, if~$\sigma$ holds for a
given~$\TT$-model~$M$ then~$\sigma$ is provable modulo~$\TT$.} In place of the
algebraic certificate we now have a logical certificate, a \emph{proof}
of~$\sigma$. However, this generalized statement is typically false, even for
the generic model~$U_\TT$: The statement
\begin{multline*}
  \qquad\qquad\qquad\Set[\TT] \models \ulcorner\text{for any geometric sequent~$\sigma$,} \\
    \text{if~$\sigma$ holds for~$U_\TT$ then~$\ul{\TT}$
    proves~$\sigma$}\urcorner\qquad\qquad\qquad
\end{multline*}
does not hold.\footnote{Here~$\ul{\TT}$ is the internal geometric theory induced
by~$\TT$, obtained by pulling back the set of sorts, the set of function
symbols and so on along the geometric morphism~$\Set[\TT] \to \Set$. For
instance, if~$\TT$ is the theory of rings, then from the internal point of view
of~$\Set[\TT]$ the theory~$\ul{\TT}$ will again be the theory of rings.
More details will be given in Section~\ref{sect:review}. The corner quotes indicate
that for sake of readability, the translation into formal language is to be
carried out by the reader.}

@@

Intuitively, the problem is that while the meaning of~$\speak{$\ul{\TT}$
proves~$\sigma$}$ is fixed, the meaning of~$\speak{$\sigma$ holds for~$U_\TT$}$ can vary
with the slice. This problem can be solved by passing from~$\ul{\TT}$ to a
varying theory, the internal theory~$\ul{\TT}/U_\TT$ defined in
Section~\ref{sect:main}. If~$\TT$ is the theory of rings, then~$\ul{\TT}/U_\TT$
is the~$\Set[\TT]$-theory of~$U_\TT$-algebras.
Unlike~$\ul{\TT}$, this theory is not the pullback of an external geometric
theory. We then have, subject to some qualifications made precise in
Section~\ref{sect:main}:

\begin{thm}\label{thm:nullstellensatz0}
Let~$\TT$ be a geometric theory. Then, internally to~$\Set[\TT]$:
\begin{equation}\tag{$\ddagger$}\label{eq:nullstellensatz}
\text{A geometric$^*$ sequent~$\sigma$ holds for~$U_\TT$ if and only
if~$\ul{\TT}/U_\TT$ proves$^*$~$\sigma$.}
\end{equation}
\end{thm}

To illustrate Theorem~\ref{thm:nullstellensatz0}, let~$\TT$ be the theory of
rings and let~$\sigma$ be of the form~$(f(x) = 0 \seq{x} g(x) = 0)$ for some
polynomials~$f$ and~$g$. To say that~$\sigma$ holds for~$U_\TT$ amounts to
saying that any zero~$x \? U_\TT$ of~$f$ is also a zero of~$g$, and to say
that~$\ul{\TT}/U_\TT$ proves~$\sigma$ amounts to saying that
in~$U_\TT[X]/(f(X))$, the free~$U_\TT$-algebra on one generator~$X$ subject to
the relation~$f(X) = 0$, the relation~$g([X]) = 0$ holds. Hence we obtain
\[ \Set[\TT] \models
  \forall f,g \? U_\TT[X]\_ \bigl(
    (\forall x \? U_\TT\_ f(x) = 0 \Rightarrow g(x) = 0) \Longleftrightarrow
      \exists h \in U_\TT[X]\_ g = hf\bigr). \]

Theorem~\ref{thm:nullstellensatz0} is a source of nongeometric sequents. Indeed,
it is the universal such source in the sense that any first-order formula
holding for~$U_\TT$ can be deduced from~\eqref{eq:nullstellensatz}:

\begin{thm}\label{thm:characterization0}
Let~$\TT$ be a geometric theory. Let~$\alpha$ be a first-order
formula over the signature of~$\TT$. Then the following statements are
equivalent.
\begin{enumerate}
\item The formula~$\alpha$ holds for~$U_\TT$.
\item The formula~$\alpha$ is provable in first-order intuitionistic logic
modulo the axioms of~$\TT$ and the additional axiom
scheme~\eqref{eq:nullstellensatz}.
\end{enumerate}
\end{thm}

\paragraph{Acknowledgments} XXX

@@

From the point of view of~$\Set[\TT]$, a model of~$\ul{\TT}/U_\TT$ is a model
of~$\ul{\TT}$ equipped with a~$\ul{\TT}$-homomorphism from~$U_\TT$. In
particular, the identity~$(U_\TT \to U_\TT)$ is a model of~$\ul{\TT}/U_\TT$.
We then have:

% We would not expect the statement
% \[ \text{``for any geometric sequent~$\sigma$, if~$\sigma$ holds for~$A$
% then the theory of rings proves~$\sigma$''} \]
% to hold for an ordinary ring~$A$ in the topos of sets.
%
% Let~$M$ be a~$\TT$-model. We then trivially have
% signature of~$\TT$. While we always have the implication
% \[ \text{$\TT$ proves~$\sigma$} \quad\Longrightarrow\quad
%   \text{$\sigma$ holds for~$M$}, \]
% the converse is of course typically false,
%
% Furthermore, in a first introduction, it is useful to
% restrict to an instructive special case, such as the theory of (commutative) rings.
%
% A simple version of the classical Nullstellensatz, valid for algebraically
% closed fields, states: Let~$f \in A[X]$ be a polynomial. If
% \[ \forall x \in A\_ (f(x) = 0 \in A \Rightarrow 1 = 0 \in A), \]
% then there is a polynomial~$g \in A[X]$ such that~$gf = 1$. This polynomial~$g$
% can be regarded as an ``algebraic certificate'' that the equation~$f(x) = 0$
% only has a solution in the trivial case~$1 = 0$. The conclusion can also be phrased as
% follows: In the free~$A$-algebra on one generator modulo the relation~$f(X) =
% 0$, the relation~$1 = 0$ holds.
%
% Let~$A$ be a specific ring. Let~$f,g \in \ZZ[X]$ be
% polynomials. While we always have the implication
% \[ g([X]) = 0 \in \ZZ[X]/(f) \quad\Longrightarrow\quad
%   \forall x \in A\_ (f(x) = 0 \Rightarrow g(x) = 0), \]
% the converse is typically false: Just that~$g(x) = 0$ holds for every zero
% of~$f$ in~$A$ does not imply that~$g([X]) = 0$ holds for the generic zero
% of~$f$ in the free ring on one generator~$X$ modulo the relation~$f(X) = 0$.
%
%
% the case that~$\TT$ is a Horn theory (where the axioms are of the
% form~$(\phi \seq{\vec x} \psi)$ with the formulas~$\phi$ and~$\psi$ being
% finite conjunctions of atomic propositions (equality and the given relations)),
% such that we can switch from a logical terminology to a more algebraic one.
% In the Horn case, for any set~$X$ and any set~$R$ of atomic propositions the
% term algebra construction yields~$\TT\langle X | R\rangle$, the
% free~$\TT$-model generated by~$X$ modulo the relations~$R$. We will also abuse
% notation to pretend that~$\TT$ is single-sorted.
%
% Let~$M$ be a specific~$\TT$-model and assume that for
% all~$n$-tuples~$x = (x_1,\ldots,x_n)$ of elements of~$M$, if some set~$R_1$ of
% atomic propositions is satisfied for~$x$ then so is some set~$R_2$. In this
% case it is unreasonable to expect that~$R_2$ holds for the~$n$-tuple of
% generators of~$\TT\langle e_1,\ldots,e_n | R_1\rangle$.
%
% let~$R_1$ and~$R_2$ be sets of atomic
% propositions with~$n$ free variables. Then
% and assume, for some sets of relations~$R_1$
% and~$R_2$
%
% The term algebra of a Horn theory~$\TT$ is always a model of~$\TT$, such that
